The relative roles of genetic, ecological and environmental factors in species radiations

### 2018-02-19 paper/draft.tex
restructure and update intro

%% [NEW ANGLE:]
% Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos \& Ricklefs 2009) %TODO: ref
Due to global change, however, island biota are also increasingly at risk.
While often featuring unique species communities and harboring countless endemics, %TODO: refs
islands are especially threatened by climate change, human impact and invasive species. %TODO: refs
This makes understanding the functioning and evolution of island communities a high priority objective.

@@

About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM). %TODO: ref
The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
whith islands emerging from sea level, increasing over time in size and topographic complexity, before eroding and subsiding in the ocean again.
In doing so, the model incorporates concepts of niche theory, speciation and species succession.
Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory.

@@

The problem of investigating these processes is the time frame needed for a proper evaluation.
Because islands biogeography takes into account evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult to obtain.
In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography
--- mechanistic (/simulation) models.
Unfortunately, most island models failed to include more explicit genetic or micro evolution processes.
% reasons: underlying theories: Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
%A subset of these questions deals with 
Instead, some models featured a concept known as ``protracted speciaton'' \cite{rosindellXXX}, where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this and genetic mechanisms have important ramifications for speciation/radiation and species richness.
[EXAMPLES!]

@@

Another shortfall of the reviewd models was the lack of integrating ecological niches and adaption.

@@

In the following, we present a model that remedies both of the adressed shortcomings by implementing genetic processes on a niche-explicit background
using an island biogeography framework.
We will adress the questions of what relative impact does (a) genetic linkage and (b) reproductive isolation have on emerging species numbers and diversity,
and how (c) landscape heterogeneity (habitat niche diversity) and (d) topographical complexity differently affect radiatons.
[further questions thus tackled: traits of colonizer species, ...]

>>>>

%% [NEW ANGLE:]
% Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos \& Ricklefs 2009) %TODO: ref
While often featuring unique species communities and harboring countless endemics, %TODO: refs
islands are especially threatened by climate change, human impact and invasive species. %TODO: refs
This makes understanding the functioning and evolution of island communities a high priority objective.

About a decade ago, island biogeography theory gained new momentum by Whittaker et al.'s General Dynamic Model (GDM). %TODO: ref
The conceptual framework traces island communities along the geomorphological trajectory of hotspot islands,
whith islands emerging from sea level, increasing over time in size and topographic complexity, before eroding and subsiding in the ocean again.
In doing so, the model incorporates concepts of niche theory, speciation and species succession.
Yet, there are still many processes and drivers deemed important, which are not considered by any formal theory and often little investigated.

@@

The problem of investigating these processes is the time frame and study extent needed for a proper evaluation.
Because islands biogeography takes into account evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult to obtain.

In a recent review \cite{leidinger2017biodiversity}, we analyzed a powerful alternative to field experiments in island biogeography: mechanistic (/simulation) models.
Although many of the reviewed models integrate multiple processes at a time, unfortunately, most models failed to include more explicit genetic or micro evolution processes.
Instead, some models used concepts such as ``protracted speciaton'' \cite{rosindellXXX},
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this and genetic mechanisms have important ramifications for speciation/radiation and species richness.

@@

@@

Another shortfall of the reviewed models was the lack of integrating ecological niches and adaption.

@@

In the following, we present a model that remedies both of the adressed shortcomings
by implementing genetic processes on a niche-explicit background
using an island biogeography framework.
We will adress the questions of what relative impact does (a) genetic linkage and
(b) reproductive isolation have on emerging species numbers and diversity,
and (c) which genetic and functional trait characteristics distinguish successful colonizer,
strong competitor and diversifying lineages.
